---
title: "12 IP weighting and marginal structural models"
author: "Bill Behrman"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Packages
library(tidyverse)

# Parameters
  # NHEFS data
file_nhefs <- here::here("data/nhefs.rds")
  # Common code
file_common <- here::here("_common.R")

# Print tibble
kable <- function(x, nsmall = 2, ...) {
  knitr::kable(x, format.args = list(digits = nsmall, nsmall = nsmall), ...)
}
# Print min, mean, and max of vector
kable_summary <- function(x, nsmall = 2, ...) {
  tibble(min = min(x), mean = mean(x), max = max(x)) %>%
    knitr::kable(format.args = list(digits = nsmall, nsmall = nsmall), ...)
}

#===============================================================================

# Run common code
source(file_common)

# NHEFS data
nhefs <- read_rds(file_nhefs)

# NHEFS censored for those with weight measurements in 1982
nhefs_censored <- 
  nhefs %>% 
  drop_na(wt82, wt82_71)
```

## 12.1 The causal question

The data are from the [National Health and Nutrition Examination Survey 1 Epidemiologic Follow-up Study](https://wwwn.cdc.gov/nchs/nhanes/nhefs/) (NHEFS). The data are drawn from two questionnaires, one taken in 1971 and the other in 1982. We will be using these variables:

* `active`: In your usual day, how active are you? in 1971
  * 0: Very active
  * 1: Moderately active
  * 2: Inactive
* `age`: Age in 1971
* `death`: Death by 1992
  * 0: No
  * 1: Yes
* `education`: Amount of education by 1971
  * 1: 8th grade or less
  * 2: High school dropout
  * 3: High school
  * 4: College dropout
  * 5: College or more
* `exercise`: In recreation, how much exercise? in 1971
  * 0: Much exercise
  * 1: Moderate exercise
  * 2: Little or no exercise
* `qsmk`: Quit smoking between 1971 and 1982 questionnaires
  * 0: No
  * 1: Yes
* `race`: Race in 1971
  * 0: White
  * 1: Black or other
* `sex`: Sex
  * 0: Male
  * 1: Female
* `smokeintensity`: Number of cigarettes smoked per day in 1971
* `smokeintensity82_71`: Increase in number of cigarettes smoked per day between 1971 and 1982
* `smokeyrs`: Years of smoking
* `wt71`: Weight in 1971 (kilograms)
* `wt82`: Weight in 1982 (kilograms)
* `wt82_71`: Weight change between 1971 and 1982 (kilograms)

The causal question is: What is the average causal effect of smoking cessation (`qsmk`) on body weight gain (`wt82_71`)?

In the following we will examine the censored dataset of those with weight measurements in 1982. Number of people in dataset.

```{r}
nrow(nhefs_censored)
```

Range of ages.

```{r}
range(nhefs_censored$age)
```

Average weight gains in quitters and non-quitters.

```{r}
v <- 
  nhefs_censored %>% 
  group_by(qsmk) %>% 
  summarize(across(wt82_71, mean))

kable(v, nsmall = 1)
```

The average weight gain was `r format(v %>% filter(qsmk == "1") %>% pull(wt82_71), digits = 1, nsmall = 1)` kg in quitters and `r format(v %>% filter(qsmk == "0") %>% pull(wt82_71), digits = 1, nsmall = 1)` kg in non-quitters.

Create a simple linear model to get an estimate the difference in average weight gains with a 95% confidence interval.

```{r}
fit <- lm(wt82_71 ~ qsmk, data = nhefs_censored)

broom::tidy(fit, conf.int = TRUE) %>% 
  filter(term == "qsmk1") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

Table 12.1.

```{r}
var_descriptions <- 
  c(
    age = "Age (years)",
    male = "Men (%)",
    white = "White (%)",
    university = "University (%)",
    wt71 = "Weight (kg)",
    smokeintensity = "Cigarettes per day",
    smokeyrs = "Years smoking",
    little_exercise = "Little exercise (%)",
    inactive = "Inactive life (%)"
  )

nhefs_censored %>% 
  transmute(
    qsmk,
    age,
    male = sex == "0",
    white = race == "0",
    university = education == "5",
    wt71,
    smokeintensity,
    smokeyrs,
    little_exercise = exercise == "2",
    inactive = active == "2"
  ) %>% 
  group_by(qsmk) %>% 
  summarize(across(everything(), mean)) %>% 
  mutate(
    across(c(male, white, university, little_exercise, inactive), ~ 100 * .)
  ) %>% 
  pivot_longer(cols = !qsmk, names_to = "Mean baseline characteristics") %>% 
  pivot_wider(names_from = qsmk, names_prefix = "A = ") %>% 
  mutate(across(`Mean baseline characteristics`, ~ var_descriptions[.])) %>% 
  relocate(`A = 0`, .after = `A = 1`) %>%
  kable(nsmall = 1)
```

## 12.2 Estimating IP weights via modeling

Fit logistic regression model for propensity scores.

```{r}
fit <- 
  glm(
    qsmk ~ sex + age + I(age^2) + race + education + wt71 + I(wt71^2) +
      smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
      active + exercise,
    family = binomial(),
    data = nhefs_censored
  )
```

Non-stabilized IP weights for treatment.

```{r}
ip_w_a <- 
  case_when(
    nhefs_censored$qsmk == "0" ~ 1 / (1 - predict(fit, type = "response")),
    nhefs_censored$qsmk == "1" ~ 1 / predict(fit, type = "response"),
    TRUE ~ NA_real_
  )

kable_summary(ip_w_a)
```

Estimate average treatment effect (ATE) using weighted least squares. This is a valid method for estimating the ATE but not its standard error.

```{r}
fit_lm <- lm(wt82_71 ~ qsmk, data = nhefs_censored, weights = ip_w_a)

broom::tidy(fit_lm)
```

Estimate ATE and its standard error using generalized estimating equation model.

```{r}
fit_geeglm <- 
  geepack::geeglm(
    wt82_71 ~ qsmk, data = nhefs_censored, weights = ip_w_a, id = seqn
  )

broom::tidy(fit_geeglm)
```

The two estimates for the ATE are very close.

```{r}
ate_lm <- coef(fit_lm)[["qsmk1"]]
ate_geeglm <- coef(fit_geeglm)[["qsmk1"]]

ate_lm - ate_geeglm
```

Estimate of ATE with 95% confidence interval.

```{r}
v <- 
  broom::tidy(fit_geeglm, conf.int = TRUE) %>% 
  filter(term == "qsmk1") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high)

kable(v, nsmall = 1)
```

```{r, echo=FALSE}
v %>% 
  save_result(
    file = "data/ate.rds",
    section = "12.2",
    method = "IP weighting - Non-stabilized",
    dataset = "Censored"
  )
```

Plot treatment levels for each confounder.

```{r}
plot_treatment <- function(var, weights, bins = 25) {
  v <- 
    nhefs_censored %>% 
    add_column(weights = weights) %>% 
    group_by({{var}}, qsmk) %>%
    summarize(n = sum(weights)) %>% 
    ungroup()
  
  if (is.factor(pull(nhefs_censored, {{var}}))) {
    v %>% 
      ggplot(aes({{var}}, n)) +
      geom_col(aes(fill = qsmk), position = "dodge")
  } else if (is.numeric(pull(nhefs_censored, {{var}}))) {
    v %>% 
      ggplot(aes({{var}})) +
      geom_freqpoly(aes(color = qsmk, weight = n), bins = bins)
  } else {
    NULL
  }
}

confounders <- 
  vars(
    sex, age, race, education, smokeintensity, smokeyrs, active, exercise, wt71
  )

confounders %>%
  map(plot_treatment, weights = ip_w_a) %>% 
  walk(print)
```

In the pseudo-population with nonstabilized IP weighting, the treatment populations are roughly equal.

## 12.3 Stabilized IP weights

Stabilized IP weights for treatment.

```{r}
ip_sw_a <- 
  case_when(
    nhefs_censored$qsmk == "0" ~ 1 - mean(nhefs_censored$qsmk == "1"),
    nhefs_censored$qsmk == "1" ~ mean(nhefs_censored$qsmk == "1"),
    TRUE ~ NA_real_
  ) * 
  ip_w_a

kable_summary(ip_sw_a)
```

Estimate ATE and its standard error using generalized estimating equation model.

```{r}
fit_geeglm <- 
  geepack::geeglm(
    wt82_71 ~ qsmk, data = nhefs_censored, weights = ip_sw_a, id = seqn
  )

broom::tidy(fit_geeglm)
```

Estimate of ATE with 95% confidence interval.

```{r}
v <- 
  broom::tidy(fit_geeglm, conf.int = TRUE) %>% 
  filter(term == "qsmk1") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high)

kable(v, nsmall = 1)
```

```{r, echo=FALSE}
v %>% 
  save_result(
    file = "data/ate.rds",
    section = "12.3",
    method = "IP weighting - Stabilized",
    dataset = "Censored"
  )
```

Plot treatment levels for each confounder.

```{r}
confounders %>%
  map(plot_treatment, weights = ip_sw_a) %>% 
  walk(print)
```

In the pseudo-population with stabilized IP weighting, the treatment populations are now unequal. These populations reflect the fact that approximately `r format(100 * mean(nhefs_censored$qsmk == "0"), digits = 1, nsmall = 1)`% of the participants did not quit smoking (`qsmk` = 0) and `r format(100 * mean(nhefs_censored$qsmk == "1"), digits = 1, nsmall = 1)`% of the participants did (`qsmk` = 1).

### Fine Point 12.2 Checking positivity

Count number of individuals in all combinations of `sex`, `age`, and `qsmk`.

```{r}
v <- 
  nhefs_censored %>% 
  expand(sex, age = full_seq(age, 1), qsmk) %>% 
  left_join(
    nhefs_censored %>% 
      mutate(n = 1),
    by = c("sex", "age", "qsmk")
  ) %>% 
  count(sex, age, qsmk, wt = n)

v
```

Number of combinations by number of individuals in combination.

```{r}
v %>% 
  ggplot(aes(n)) +
  geom_bar() +
  labs(
    title = "Number of combinations by number of individuals in combination",
    x = "Number of individuals in combination",
    y = "Number of combinations"
  )
```

Eleven combinations have no individuals.

```{r}
v %>%
  filter(n == 0) %>% 
  kable(nsmall = 0)
```

## 12.4 Marginal structural models

In this section, we will estimate causal effect of the change in smoking intensity (`smkintensity82_71`), a continuous variable, on the average weight gain (`wt82_71`).

Distribution of `smokeintensity`.

```{r}
nhefs_censored %>% 
  ggplot(aes(smokeintensity)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  labs(title = "Distribution of smokeintensity")
```

Create subset of data with individuals who smoked 25 or fewer cigarettes per day at baseline.

```{r}
nhefs_censored_smoke_25 <- 
  nhefs_censored %>% 
  filter(smokeintensity <= 25)
```

Number of individuals in subset.

```{r}
nrow(nhefs_censored_smoke_25)
```

Numerator for IP weights.

```{r}
treatment <- nhefs_censored_smoke_25$smkintensity82_71

ip_numerator <- 
  dnorm(
    treatment,
    mean = mean(treatment),
    sd = sd(treatment)
  )
```

The numerator for the IP weights approximates `smkintensity82_71` with a normal distribution. Here is the actual distribution of `smkintensity82_71` with its normal approximation.

```{r}
normal_approx <-
  tibble(
    x = seq(min(treatment), max(treatment), length.out = 201),
    y = dnorm(x, mean = mean(treatment), sd = sd(treatment))
  )

nhefs_censored_smoke_25 %>% 
  ggplot() +
  geom_histogram(
    aes(smkintensity82_71, stat(density)),
    binwidth = 5,
    boundary = 0
  ) +
  geom_line(aes(x, y), data = normal_approx, color = "red") +
  labs(
    title = "Distribution of smokeintensity82_71",
    subtitle = "With normal approximation in red"
  )
```

Denominator for IP weights.

```{r}
fit <- 
  lm(
    smkintensity82_71 ~ sex + age + I(age^2) + race + education +
      wt71 + I(wt71^2) + smokeintensity + I(smokeintensity^2) +
      smokeyrs + I(smokeyrs^2) + active + exercise,
    data = nhefs_censored_smoke_25
  )

ip_denominator <- 
  dnorm(
    treatment,
    mean = predict(fit, type = "response"),
    sd = sigma(fit)
  )
```

IP weights for marginal structural model.

```{r}
ip_msm <- ip_numerator / ip_denominator

kable_summary(ip_msm)
```

Fit marginal structural model.

```{r}
fit <- 
  geepack::geeglm(
    wt82_71 ~ smkintensity82_71 + I(smkintensity82_71^2),
    data = nhefs_censored_smoke_25,
    weights = ip_msm,
    id = seqn
  )

broom::tidy(fit)
```

Coefficients of model.

```{r}
broom::tidy(fit) %>% 
  select(term, estimate) %>% 
  kable(nsmall = 1)
```

Mean weight gain with 95% confidence interval for constant smoking intensity.

```{r}
broom::tidy(fit, conf.int = TRUE) %>% 
  filter(term == "(Intercept)") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

Mean weight gain with 95% confidence interval for increase in smoking intensity of 20 cigarettes per day.

```{r}
fit <- 
  geepack::geeglm(
    wt82_71 ~ I(smkintensity82_71 - 20) + I((smkintensity82_71 - 20)^2),
    data = nhefs_censored_smoke_25,
    weights = ip_msm,
    id = seqn
  )

broom::tidy(fit, conf.int = TRUE) %>% 
  filter(term == "(Intercept)") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

Marginal structural model for causal effect of quitting smoking (`qsmk`) on the risk of death by 1992 (`death`). For this calculation, we will convert the type of `qsmk` and `death` from factor to double.

```{r, warning=FALSE}
fit <- 
  nhefs_censored %>% 
  mutate(across(c(qsmk, death), ~ as.double(.) - 1)) %>%
  geepack::geeglm(
    death ~ qsmk, family = binomial(), data = ., weights = ip_sw_a, id = seqn
  )

broom::tidy(fit)
```

Estimate of causal odds ratio with 95% confidence interval.

```{r}
broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE) %>% 
  filter(term == "qsmk") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

## 12.5 Effect modification and marginal structural models

IP weights.

```{r}
fit <- glm(qsmk ~ sex, family = binomial(), data = nhefs_censored)

ip_sw_a <- 
  case_when(
    nhefs_censored$qsmk == "0" ~ (1 - predict(fit, type = "response")),
    nhefs_censored$qsmk == "1" ~ predict(fit, type = "response"),
    TRUE ~ NA_real_
  ) *
  ip_w_a

kable_summary(ip_sw_a)
```

Fit marginal structural model.

```{r}
fit <- 
  geepack::geeglm(
    wt82_71 ~ qsmk * sex, data = nhefs_censored, weights = ip_sw_a, id = seqn
  )

broom::tidy(fit)
```

Estimate of effect modification by sex.

```{r}
broom::tidy(fit, conf.int = TRUE) %>% 
  filter(term == "qsmk1:sex1") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

## 12.6 Censoring and missing data

Number of individuals in full dataset.

```{r}
nrow(nhefs)
```

Number of individual whose body weight in 1982 is missing.

```{r}
sum(is.na(nhefs$wt82))
```

Add `censored` variable to `nhefs` to indicate whether or not body weight in 1982 (`wt82`) is missing.

```{r}
nhefs <- 
  nhefs %>% 
  mutate(censored = if_else(!is.na(wt82), "0", "1") %>% as.factor())
```

The treatment `qsmk` is associated with censoring.

```{r}
v <- 
  nhefs %>% 
  group_by(qsmk) %>% 
  summarize(censored = mean(censored == "1"))

kable(v)
```

`r format(100 * (v %>% filter(qsmk == "1") %>% pull(censored)), digits = 2)`% of quitters versus `r format(100 * (v %>% filter(qsmk == "0") %>% pull(censored)), digits = 2)`% of non-quitters were censored.

The predictor `wt71` is also associated with censoring.

```{r}
v <- 
  nhefs %>% 
  group_by(censored) %>% 
  summarize(across(wt71, mean))

kable(v, nsmall = 1)
```

The average baseline weight was `r format(v %>% filter(censored == "1") %>% pull(wt71), digits = 3)` kg in the censored versus `r format(v %>% filter(censored == "0") %>% pull(wt71), digits = 3)` kg in the uncensored.

Stabilized IP weights for treatment.

```{r}
ip_numerator <- 
  case_when(
    nhefs$qsmk == "0" ~ 1 - mean(nhefs$qsmk == "1"),
    nhefs$qsmk == "1" ~ mean(nhefs$qsmk == "1"),
    TRUE ~ NA_real_
  )

fit_denominator <- 
  glm(
    qsmk ~ sex + age + I(age^2) + race + education + wt71 + I(wt71^2) +
      smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
      active + exercise,
    family = binomial(),
    data = nhefs
  )

ip_denominator <- 
  case_when(
    nhefs$qsmk == "0" ~ 1 - predict(fit_denominator, type = "response"),
    nhefs$qsmk == "1" ~ predict(fit_denominator, type = "response"),
    TRUE ~ NA_real_
  )

ip_sw_a <- ip_numerator / ip_denominator 
```

Stabilized IP weights for censoring.

```{r}
fit_numerator <- glm(censored ~ qsmk, family = binomial(), data = nhefs)

ip_numerator <- 
  case_when(
    nhefs$censored == "0" ~ 1 - predict(fit_numerator, type = "response"),
    nhefs$censored == "1" ~ predict(fit_numerator, type = "response"),
    TRUE ~ NA_real_
  )

fit_denominator <- 
  glm(
    censored ~ qsmk + sex + age + I(age^2) + race + education +
      wt71 + I(wt71^2) + smokeintensity + I(smokeintensity^2) +
      smokeyrs + I(smokeyrs^2) + active + exercise,
    family = binomial(),
    data = nhefs
  )

ip_denominator <- 
  case_when(
    nhefs$censored == "0" ~ 1 - predict(fit_denominator, type = "response"),
    nhefs$censored == "1" ~ predict(fit_denominator, type = "response"),
    TRUE ~ NA_real_
  )

ip_sw_c <- ip_numerator / ip_denominator
```

Stabilized IP weights for treatment and censoring

```{r}
ip_sw_ac = ip_sw_a * ip_sw_c

kable_summary(ip_sw_ac[nhefs$censored == "0"])
```

Estimate ATE and its standard error using generalized estimating equation model.

```{r}
fit_geeglm <- 
  geepack::geeglm(
    wt82_71 ~ qsmk,
    data = nhefs_censored,
    weights = ip_sw_ac[nhefs$censored == "0"],
    id = seqn
  )

broom::tidy(fit_geeglm)
```

Estimate of ATE with 95% confidence interval.

```{r}
v <- 
  broom::tidy(fit_geeglm, conf.int = TRUE) %>% 
  filter(term == "qsmk1") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high)

kable(v, nsmall = 1)
```

```{r, echo=FALSE}
v %>% 
  save_result(
    file = "data/ate.rds",
    section = "12.6",
    method = "IP weighting - Stabilized",
    dataset = "Full"
  )
```

