
# 17 Causal survival analysis {-}

```{r message=FALSE, warning=FALSE}
# Packages
library(tidyverse)

# Parameters
  # NHEFS data
file_nhefs <- here::here("data/nhefs.rds")

# Round and format vector
round_format <- function(x, nsmall = 2, ...) {
  format(round(x, digits = nsmall), nsmall = nsmall, ...)
}
# Print tibble
kable <- function(x, cols = where(is.double), nsmall = 2, align = "r", ...) {
  x %>%
    mutate(across({{cols}}, round_format, nsmall = nsmall)) %>%
    knitr::kable(align = align, ...) %>% 
    kableExtra::kable_styling(full_width = FALSE, position = "left")
}
# Print min, mean, and max of vector
kable_summary <- function(x, nsmall = 2, ...) {
  tibble(min = min(x), mean = mean(x), max = max(x)) %>%
    kable(nsmall = nsmall, ...)
}

#===============================================================================

# NHEFS data with qsmk as a double
nhefs <- 
  read_rds(file_nhefs) %>% 
  mutate(qsmk = as.double(qsmk) - 1)
```

## 17.1 Hazards and risks {-}

In the following, we will use these variables:

* `dadth`: Day of death
* `death`: Death by 1992
  * 0: No
  * 1: Yes
* `modth`: Month of death
* `yrdth`: Year of death

Number of people in dataset.

```{r}
nrow(nhefs)
```

Range of ages.

```{r}
range(nhefs$age)
```

The number of people who died by the end of 1992.

```{r}
v <- 
  nhefs %>% 
  count(death)

kable(v)
```

Of the `r nrow(nhefs)` individuals in the dataset, `r v %>% filter(death == "1") %>% pull(n)` died before the end of 1992, and the remaining `r v %>% filter(death == "0") %>% pull(n)` survived.

The variables `death`, `yrdth`, `modth`, and `dadth`.

```{r}
nhefs %>% 
  count(death, !is.na(yrdth), !is.na(modth), !is.na(dadth))
```

Note that four individuals have months and days of death but no year. They are classified by the `death` variable as having not died. For the following, we will assume that the `death` variable is correct.

The follow-up period began on January 1, 1983 and lasted through December 31, 1992. For those who died, we will add the variable `time_death` for the month of the follow-up in which they died, ranging from 1 for January 1983 to 120 for December 1992.

```{r}
nhefs <- 
  nhefs %>% 
  mutate(
    time_death = if_else(death == "1", 12 * (yrdth - 83) + modth, NA_real_)
  )
```

Range of values for `time_death`.

```{r}
range(nhefs$time_death, na.rm = TRUE)
```

Number of individuals who died by whether or not they quit smoking (`qsmk`).

```{r}
v <- 
  nhefs %>% 
  filter(!is.na(time_death)) %>% 
  count(qsmk)

kable(cols = NULL, v)
```

Of the `r sum(v$n)` individuals who died, `r v %>% filter(qsmk == 0) %>% pull(n)` did not quit smoking and `r v %>% filter(qsmk == 1) %>% pull(n)` did quit smoking.

For each treatment group (`qsmk`) and each follow-up month, calculate the cumulative deaths (`deaths_cum`), the number who survived (`survived`), and the proportion who survived (`survival`).

```{r}
time_max <- max(nhefs$time_death, na.rm = TRUE)

survival_1 <- 
  expand_grid(qsmk = 0:1, time = 0:time_max) %>% 
  left_join(
    nhefs %>% 
      filter(death == "1") %>% 
      count(qsmk, time = time_death, name = "deaths"),
    by = c("qsmk", "time")
  ) %>% 
  replace_na(list(deaths = 0)) %>% 
  group_by(qsmk) %>% 
  mutate(
    deaths_cum = cumsum(deaths),
    survived = sum(nhefs$qsmk == first(qsmk)) - deaths_cum,
    survival = survived / sum(nhefs$qsmk == first(qsmk))
  ) %>% 
  ungroup()

survival_1
```

Figure 17.1.

```{r}
survival_plot_details <- function(y_quitters = 0.74) {
  list(
    annotate(
      "text",
      x = 119,
      y = c(0.87, y_quitters),
      hjust = 1,
      label = c("Non-quitters", "Quitters")
    ),
    coord_cartesian(ylim = c(0.5, 1)),
    scale_x_continuous(breaks = scales::breaks_width(20)),
    theme(legend.position = "none"),
    labs(
      x = "Month of follow-up",
      y = "Survival probability"
    )
  )
}

survival_1 %>% 
  ggplot(aes(time, survival, color = as.factor(qsmk))) +
  geom_line() +
  survival_plot_details() +
  labs(title = "Figure 17.1")
```

Log-rank test to compare survival curves.

```{r}
test <- 
  nhefs %>% 
  transmute(
    qsmk,
    time = replace_na(time_death, replace = time_max),
    event = as.double(death) - 1
  )  %>% 
  survival::survdiff(
    survival::Surv(time = time, event = event) ~ qsmk,
    data = .
  )

broom::glance(test) %>% 
  select(p_value = p.value) %>% 
  kable(nsmall = 3)
```

Survival at 120 months.

```{r}
v <- 
  survival_1 %>% 
  filter(time == 120) %>% 
  select(qsmk, survival)

v %>% 
  kable(cols = survival, nsmall = 3)
```

Survival at 120 months was `r round_format(100 * (v %>% filter(qsmk == "0") %>% pull(survival)), nsmall = 1)`% among non-quitters and `r round_format(100 * (v %>% filter(qsmk == "1") %>% pull(survival)), nsmall = 1)`% among quitters. The risk at 120 months was `r round_format(100 * (1 - (v %>% filter(qsmk == "0") %>% pull(survival))), nsmall = 1)`% among non-quitters and `r round_format(100 * (1 - (v %>% filter(qsmk == "1") %>% pull(survival))), nsmall = 1)`% among quitters.

Hazard at 120 months.

```{r}
v <- 
  survival_1 %>% 
  mutate(
    survived_prev = lag(survived),
    hazard = deaths / survived_prev
  ) %>% 
  filter(time == 120) %>% 
  select(qsmk, time, deaths, survived_prev, hazard)

v %>% 
  kable(cols = hazard, nsmall = 4)
```

The hazard at 120 months was `r round_format(100 * (v %>% filter(qsmk == "0") %>% pull(hazard)), nsmall = 2)`% among non-quitters and `r 100 * (v %>% filter(qsmk == "1") %>% pull(hazard))`% among quitters.

## 17.2 From hazards to risks {-}

Time-event tibble for person in person-time format.

```{r}
time_event <- function(death, time_death) {
  stopifnot(death %in% c("0", "1"))
  if (death == "0") {
    tibble(time = seq_len(time_max) - 1, event = 0)
  } else if (death == "1") {
    tibble(
      time = seq_len(time_death) - 1,
      event = rep(0:1, c(time_death - 1, 1))
    )
  }
}
```

NHEFS dataset in person-time format.

```{r}
nhefs_pt <- 
  nhefs %>% 
  select(seqn, qsmk, death, time_death) %>% 
  rowwise() %>% 
  mutate(time_event = list(time_event(death, time_death))) %>% 
  unnest(time_event)
```

Number of rows in `nhefs_pt`.

```{r}
nrow(nhefs_pt)
```

Fit logistic regression for hazards.

```{r}
fit <- 
  glm(
    event ~ poly(time, 2) + qsmk * poly(time, 2),
    family = binomial(),
    data = nhefs_pt
  )

broom::tidy(fit)
```

Calculate survival curves.

```{r}
survival <- function(fit) {
  expand_grid(qsmk = 0:1, time = 0:time_max) %>% 
    group_by(qsmk) %>% 
    mutate(
      survival = 
        cumprod(
          1 - predict(fit, newdata = tibble(qsmk, time), type = "response")
        ) %>% 
        lag(default = 1)
    ) %>% 
    ungroup()
}
```

Figure 17.4.

```{r}
survival_2 <- survival(fit)

survival_2 %>% 
  ggplot(aes(time, survival, color = as.factor(qsmk))) +
  geom_line() +
  survival_plot_details() +
  labs(title = "Figure 17.4")
```

Superposition of Figures 17.1 and 17.4.

```{r}
ggplot(mapping = aes(time, survival)) +
  geom_line(aes(group = qsmk), data = survival_1, color = "grey60") +
  geom_line(aes(color = as.factor(qsmk)), data = survival_2) +
  survival_plot_details() + 
  labs(title = "Superposition of Figures 17.1 and 17.4")
```

## 17.4 IP weighting of marginal structural models {-}

Stabilized IP weights for treatment.

```{r}
ip_numerator <- 
  case_when(
    nhefs$qsmk == 0 ~ 1 - mean(nhefs$qsmk),
    nhefs$qsmk == 1 ~ mean(nhefs$qsmk),
    TRUE ~ NA_real_
  )

fit_denominator <- 
  glm(
    qsmk ~ sex + poly(age, 2) + race + education + poly(wt71, 2) +
      poly(smokeintensity, 2) + poly(smokeyrs, 2) + active + exercise,
    family = binomial(),
    data = nhefs
  )

ip_denominator <- 
  case_when(
    nhefs$qsmk == 0 ~ 1 - predict(fit_denominator, type = "response"),
    nhefs$qsmk == 1 ~ predict(fit_denominator, type = "response"),
    TRUE ~ NA_real_
  )

ip_sw_a <- ip_numerator / ip_denominator

kable_summary(ip_sw_a)
```

NHEFS dataset with weights in person-time format.

```{r}
nhefs_pt <- 
  nhefs %>% 
  transmute(seqn, qsmk, weight = ip_sw_a, death, time_death) %>% 
  rowwise() %>% 
  mutate(time_event = list(time_event(death, time_death))) %>% 
  unnest(time_event)
```

Fit logistic regression for hazards.

```{r, warning=FALSE}
fit <- 
  glm(
    event ~ poly(time, 2) + qsmk * poly(time, 2),
    family = binomial(),
    data = nhefs_pt,
    weights = weight
  )

broom::tidy(fit)
```

Figure 17.6.

```{r}
survival_3 <- survival(fit)

survival_3 %>% 
  ggplot(aes(time, survival, color = as.factor(qsmk))) +
  geom_line() +
  survival_plot_details(y_quitters = 0.78) +
  labs(title = "Figure 17.6")
```

Superposition of Figures 17.4 and 17.6.

```{r}
ggplot(mapping = aes(time, survival)) +
  geom_line(aes(group = qsmk), data = survival_2, color = "grey60") +
  geom_line(aes(color = as.factor(qsmk)), data = survival_3) +
  survival_plot_details() + 
  labs(title = "Superposition of Figures 17.4 and 17.6")
```

Survival at 120 months.

```{r}
v <- 
  survival_3 %>% 
  filter(time == 120) %>% 
  select(qsmk, survival)

v %>% 
  kable(cols = survival, nsmall = 3)
```

Survival at 120 months was `r round_format(100 * (v %>% filter(qsmk == 0) %>% pull(survival)), nsmall = 1)`% among non-quitters and `r round_format(100 * (v %>% filter(qsmk == 1) %>% pull(survival)), nsmall = 1)`% among quitters for a difference of `r round_format(100 * ((v %>% filter(qsmk == 1) %>% pull(survival)) - (v %>% filter(qsmk == 0) %>% pull(survival))), nsmall = 1)`%.

Calculate survival difference at 120 months and the largest absolute value of the differences for all months. If `boot = TRUE` (the default), use bootstrap sample of data. If `boot = FALSE`, use full dataset.

```{r}
survival_diff <- function(boot = TRUE) {
  data <-
    nhefs %>%
    transmute(seqn, qsmk, weight = ip_sw_a, death, time_death) %>%
    {
      if (isTRUE(boot)) {
        slice_sample(., prop = 1, replace = TRUE)
      } else {
        .
      }
    } %>%
    rowwise() %>%
    mutate(time_event = list(time_event(death, time_death))) %>%
    unnest(time_event)

  fit <-
    glm(
      event ~ poly(time, 2) + qsmk * poly(time, 2),
      family = binomial(),
      data = data,
      weights = weight
    )

  survival(fit) %>% 
    group_by(time) %>%
    summarize(diff = survival[qsmk == 1] - survival[qsmk == 0]) %>%
    summarize(
      last = last(diff),
      max = max(abs(diff))
    )
}
```

Perform bootstrap resampling.

```{r, warning=FALSE}
set.seed(231)

n_boot <- 500

boot_out <-
  seq_len(n_boot) %>%
  map_dfr(~ survival_diff())
```

Survival difference at 120 months with 95% confidence interval calculated using bootstrap percentile method.

```{r, warning=FALSE}
v <-
  tibble(
    estimate = survival_diff(boot = FALSE)$last,
    conf_low = quantile(boot_out$last, probs = 0.025),
    conf_high = quantile(boot_out$last, probs = 0.975)
  )

kable(v, nsmall = 3)
```

The survival difference at 120 months was `r round_format(100 * v$estimate, nsmall = 1)`% with a 95% confidence interval from `r round_format(100 * v$conf_low, nsmall = 1)`% to `r round_format(100 * v$conf_high, nsmall = 1)`% based on `r n_boot` bootstrap samples.

The largest absolute value of the differences for all months with 95% confidence interval calculated using bootstrap percentile method. Note that this a different measure than the one used in the book.

```{r, warning=FALSE}
v <-
  tibble(
    estimate = survival_diff(boot = FALSE)$max,
    conf_low = quantile(boot_out$max, probs = 0.025),
    conf_high = quantile(boot_out$max, probs = 0.975)
  )

kable(v, nsmall = 3)
```

The largest absolute value of the differences for all months was `r round_format(100 * v$estimate, nsmall = 1)`% with a 95% confidence interval from `r round_format(100 * v$conf_low, nsmall = 1)`% to `r round_format(100 * v$conf_high, nsmall = 1)`% based on `r n_boot` bootstrap samples.

## 17.5 The parametric g-formula {-}

NHEFS dataset in person-time format.

```{r}
nhefs_pt <-
  nhefs %>%
  select(
    qsmk, death, sex, age, race, education, wt71, smokeintensity,
    smkintensity82_71, smokeyrs, active, exercise, time_death
  ) %>%
  rowwise() %>%
  mutate(time_event = list(time_event(death, time_death))) %>%
  unnest(time_event)
```

Fit logistic regression for hazards.

```{r}
fit <-
  glm(
    event ~ poly(time, 2) + qsmk * poly(time, 2) + sex + poly(age, 2) + race +
      education + poly(wt71, 2) + poly(smokeintensity, 2) + smkintensity82_71 +
      poly(smokeyrs, 2) + active + exercise,
    family = binomial(),
    data = nhefs_pt
  )

broom::tidy(fit)
```

Calculate survival curve for individual.

```{r}
time_survival <- function(fit, data, qsmk) {
  tibble(
    time = 0:time_max,
    survival =
      cumprod(
        1 - predict(fit, newdata = tibble(data, qsmk, time), type = "response")
      ) %>%
      lag(default = 1)
  )
}
```

Calculate survival curves for each individual and then average.

```{r}
survival_4 <-
  nhefs %>%
  select(
    seqn, sex, age, race, education, wt71, smokeintensity, smkintensity82_71,
    smokeyrs, active, exercise
  ) %>%
  expand_grid(qsmk = 0:1) %>%
  nest(data = !c(seqn, qsmk)) %>%
  rowwise() %>%
  transmute(
    qsmk,
    time_survival = list(time_survival(fit = fit, data = data, qsmk = qsmk))
  ) %>%
  unnest(time_survival) %>%
  group_by(qsmk, time) %>%
  summarize(across(survival, mean)) %>%
  ungroup()
```

Figure 17.7.

```{r}
survival_4 %>%
  ggplot(aes(time, survival, color = as.factor(qsmk))) +
  geom_line() +
  survival_plot_details(y_quitters = 0.78) +
  labs(title = "Figure 17.7")
```

Superposition of Figures 17.6 and 17.7.

```{r}
ggplot(mapping = aes(time, survival)) +
  geom_line(aes(group = qsmk), data = survival_3, color = "grey60") +
  geom_line(aes(color = as.factor(qsmk)), data = survival_4) +
  survival_plot_details(y_quitters = 0.78) +
  labs(title = "Superposition of Figures 17.6 and 17.7")
```

The survival curves from the two figures are coincident.

Survival at 120 months.

```{r}
v <-
  survival_4 %>%
  filter(time == 120) %>%
  select(qsmk, survival)

v %>%
  kable(cols = survival, nsmall = 3)
```

Survival at 120 months was `r round_format(100 * (v %>% filter(qsmk == 0) %>% pull(survival)), nsmall = 1)`% among non-quitters and `r round_format(100 * (v %>% filter(qsmk == 1) %>% pull(survival)), nsmall = 1)`% among quitters for a difference of `r round_format(100 * ((v %>% filter(qsmk == 1) %>% pull(survival)) - (v %>% filter(qsmk == 0) %>% pull(survival))), nsmall = 1)`% (rounded from `r round_format(100 * ((v %>% filter(qsmk == 1) %>% pull(survival)) - (v %>% filter(qsmk == 0) %>% pull(survival))), nsmall = 3)`%).

Calculate survival difference at 120 months and the largest absolute value of the differences for all months. If `boot = TRUE` (the default), use bootstrap sample of data. If `boot = FALSE`, use full dataset.

```{r}
survival_diff <- function(boot = TRUE) {
  v <-
    nhefs %>%
    select(
      qsmk, death, sex, age, race, education, wt71, smokeintensity,
      smkintensity82_71, smokeyrs, active, exercise, time_death
    ) %>%
    {
      if (isTRUE(boot)) {
        slice_sample(., prop = 1, replace = TRUE)
      } else {
        .
      }
    } %>%
    mutate(row = row_number())

  data <-
    v %>%
    rowwise() %>%
    mutate(time_event = list(time_event(death, time_death))) %>%
    unnest(time_event)

  fit <-
    glm(
      event ~ poly(time, 2) + qsmk * poly(time, 2) + sex + poly(age, 2) + race +
        education + poly(wt71, 2) + poly(smokeintensity, 2) +
        smkintensity82_71 + poly(smokeyrs, 2) + active + exercise,
      family = binomial(),
      data = data
    )

  v %>%
    select(!c(qsmk, death, time_death)) %>%
    expand_grid(qsmk = 0:1) %>%
    nest(data = !c(row, qsmk)) %>%
    rowwise() %>%
    transmute(
      qsmk,
      time_survival = list(time_survival(fit = fit, data = data, qsmk = qsmk))
    ) %>%
    unnest(time_survival) %>%
    group_by(qsmk, time) %>%
    summarize(across(survival, mean)) %>%
    ungroup() %>%
    group_by(time) %>%
    summarize(diff = survival[qsmk == 1] - survival[qsmk == 0]) %>%
    summarize(
      last = last(diff),
      max = max(abs(diff))
    )
}
```

Perform bootstrap resampling.

```{r}
set.seed(231)

n_boot <- 100

boot_out <-
  seq_len(n_boot) %>%
  map_dfr(~ survival_diff())
```

Survival difference at 120 months with 95% confidence interval calculated using bootstrap percentile method.

```{r}
v <-
  tibble(
    estimate = survival_diff(boot = FALSE)$last,
    conf_low = quantile(boot_out$last, probs = 0.025),
    conf_high = quantile(boot_out$last, probs = 0.975)
  )

kable(v, nsmall = 3)
```

The survival difference at 120 months was `r round_format(100 * v$estimate, nsmall = 1)`% with a 95% confidence interval from `r round_format(100 * v$conf_low, nsmall = 1)`% to `r round_format(100 * v$conf_high, nsmall = 1)`% based on `r n_boot` bootstrap samples.

The largest absolute value of the differences for all months with 95% confidence interval calculated using bootstrap percentile method. Note that this a different measure than the one used in the book.

```{r}
v <-
  tibble(
    estimate = survival_diff(boot = FALSE)$max,
    conf_low = quantile(boot_out$max, probs = 0.025),
    conf_high = quantile(boot_out$max, probs = 0.975)
  )

kable(v, nsmall = 3)
```

The largest absolute value of the differences for all months was `r round_format(100 * v$estimate, nsmall = 1)`% with a 95% confidence interval from `r round_format(100 * v$conf_low, nsmall = 1)`% to `r round_format(100 * v$conf_high, nsmall = 1)`% based on `r n_boot` bootstrap samples.

## 17.6 G-estimation of structural nested models {-}

Fit logistic regression for treatment.

```{r}
fit <- 
  glm(
    qsmk ~ sex + poly(age, 2) + race + education + poly(wt71, 2) +
      poly(smokeintensity, 2) + poly(smokeyrs, 2) + active + exercise,
    family = binomial(),
    data = nhefs
  )
```

Predict treatment and restrict to individuals who died during follow-up.

```{r}
nhefs_died <- 
  nhefs %>% 
  mutate(qsmk_pred = predict(fit, type = "response")) %>% 
  drop_na(time_death) %>% 
  select(qsmk, time_death, qsmk_pred)
```

For potential counterfactual (`psi`), calculate score test statistic.

```{r}
g_est <- function(psi) {
  nhefs_died %>% 
    mutate(
      delta =
        (qsmk == 0 & time_death * exp(-psi) <= time_max) |
        (qsmk == 1 & time_death * exp(psi) <= time_max),
      x = delta * (qsmk - qsmk_pred)
    ) %>% 
    summarize(u = sum(x)^2 / ((n() - 1) * var(x))) %>%  
    pull(u)
}
```

95% quantile of chi-squared distribution with 1 degree of freedom.

```{r}
q_0.95 <- qchisq(p = 0.95, df = 1)

q_0.95
```

Score test statistic as a function of potential counterfactual.

```{r}
v <- 
  tibble(
    psi = seq(-0.3, 0.4, 0.001),
    u = map_dbl(psi, g_est)
  )

v %>% 
  ggplot(aes(psi, u)) +
  geom_line() +
  geom_hline(yintercept = q_0.95, color = "red") +
  scale_x_continuous(breaks = scales::breaks_width(0.1)) +
  labs(
    title = "Score test statistic as a function of potential counterfactual",
    subtitle = 
      "Red line indicates 95% quantile of chi-squared distribution with 1 degree of freedom",
    x = "Potential counterfactual",
    y = "Score test statistic"
  )
```

The plot appears to be step function, perhaps as a result of individuals being censored depending upon the value of `psi`.

The minimum of the plot.

```{r}
v %>% 
  filter(u == min(u))
```

All of the values of `psi` from `r round_format(v %>% filter(u == min(u)) %>% pull(psi) %>% min(), nsmall = 3)` to `r round_format(v %>% filter(u == min(u)) %>% pull(psi) %>% max(), nsmall = 3)` have the same minimum value for the score test statistic. We'll choose as our estimate the middle of this range.

```{r}
estimate <- 
  v %>% 
  filter(u == min(u)) %>% 
  summarize(estimate = mean(range(psi)))
```

Calculate 95% confidence interval by finding points where plot crosses the red line.

```{r}
conf_low <- 
  tibble(
    psi = seq(-0.25, -0.2, 0.0001),
    u = map_dbl(psi, g_est)
  ) %>% 
  filter((u >= q_0.95 & lead(u) < q_0.95) | (u < q_0.95 & lag(u) >= q_0.95)) %>% 
  summarize(conf_low = mean(range(psi)))

conf_high <- 
  tibble(
    psi = seq(0.3, 0.35, 0.0001),
    u = map_dbl(psi, g_est)
  ) %>% 
  filter((u < q_0.95 & lead(u) >= q_0.95) | (u >= q_0.95 & lag(u) < q_0.95)) %>% 
  summarize(conf_high = mean(range(psi)))
```

Estimate of `psi` with 95% confidence interval.

```{r}
bind_cols(estimate, conf_low, conf_high) %>% 
  kable(nsmall = 3)
```

