---
title: "13 Standardization and the parametric g-formula"
author: "Bill Behrman"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Packages
library(tidyverse)

# Parameters
  # NHEFS data
file_nhefs <- here::here("data/nhefs.rds")
  # Common code
file_common <- here::here("_common.R")

# Print tibble
kable <- function(x, nsmall = 2, ...) {
  knitr::kable(x, format.args = list(digits = nsmall, nsmall = nsmall), ...)
}
# Print min, mean, and max of vector
kable_summary <- function(x, nsmall = 2, ...) {
  tibble(min = min(x), mean = mean(x), max = max(x)) %>%
    knitr::kable(format.args = list(digits = nsmall, nsmall = nsmall), ...)
}

#===============================================================================

# Run common code
source(file_common)

# NHEFS data
nhefs <- read_rds(file_nhefs)

# NHEFS censored for those with weight measurements in 1982
nhefs_censored <- 
  nhefs %>% 
  drop_na(wt82, wt82_71)
```

## 13.1 Standardization as an alternative to IP weighting

Number of individuals in full dataset

```{r}
nrow(nhefs)
```

Number of individuals with weight measurements in 1982.

```{r}
nrow(nhefs_censored)
```

## 13.2 Estimating the mean outcome via modeling

In this section, we will use the censored dataset of those with weight measurements in 1982

Count untreated and treated individuals.

```{r}
v <- 
  nhefs_censored %>% 
  count(qsmk)

knitr::kable(v)
```

There are `r v$n[v$qsmk == "0"]` untreated individuals and `r v$n[v$qsmk == "1"]` treated individuals.

Fit linear regression model for weight gain.

```{r}
fit <- 
  lm(
    wt82_71 ~ qsmk * smokeintensity + sex + age + I(age^2) + race + education +
      wt71 + I(wt71^2) + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
      active + exercise,
    data = nhefs_censored
  )

broom::tidy(fit)
```

Predict weight gain for individuals.

```{r}
nhefs_censored_pred <- 
  nhefs_censored %>% 
  mutate(.pred = predict(fit))
```

The individual with the unique identifier 24770.

```{r}
v <- 
  nhefs_censored_pred %>% 
  filter(seqn == 24770)
```

This person's characteristics.

```{r}
v %>% 
  select(
    seqn, qsmk, sex, race, age, education, smokeintensity, smokeyrs, exercise,
    active, wt71
  ) %>% 
  knitr::kable()
```

This person was a white male who did not quit smoking. In 1971, he was age 26, a college dropout, smoked 15 cigarettes a day, had been smoking for 12 years, had moderate exercise, was very active, and weighed `r format(v$wt71, digits = 2, nsmall = 2)` kg.

This person's observed and predicted weight gain.

```{r}
v %>% 
  select(wt82_71, .pred) %>% 
  kable()
```

The person was observed to gain `r format(v$wt82_71, digits = 2, nsmall = 2)` kg between 1971 and 1982 and was predicted to gain `r format(v$.pred, digits = 2, nsmall = 2)` kg.

Observed weight gain for individuals.

```{r}
kable_summary(nhefs_censored$wt82_71, nsmall = 1)
```

Predicted weight gain for individuals.

```{r}
kable_summary(nhefs_censored_pred$.pred, nsmall = 1)
```

The mean observed weight gain and the mean predicted weight gain are very close.

```{r}
mean(nhefs_censored$wt82_71) - mean(nhefs_censored_pred$.pred)
```

Prediction residuals.

```{r}
nhefs_censored_pred %>% 
  ggplot(aes(.pred, wt82_71 - .pred)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "white", size = 2) +
  geom_smooth(method = "loess", formula = y ~ x) +
  labs(
    title = "Prediction residuals",
    x = "Predicted weight gain (kg)",
    y = "Residual (kg)"
  )
```

## 13.3 Standardizing the mean outcome to the confounder distribution

Data from Table 2.2.

```{r}
df <- 
  tribble(
    ~name,       ~L, ~A, ~Y,
    "Rheia",      0,  0,  0,
    "Kronos",     0,  0,  1,
    "Demeter",    0,  0,  0,
    "Hades",      0,  0,  0,
    "Hestia",     0,  1,  0,
    "Poseidon",   0,  1,  0,
    "Hera",       0,  1,  0,
    "Zeus",       0,  1,  1,
    "Artemis",    1,  0,  1,
    "Apollo",     1,  0,  1,
    "Leto",       1,  0,  0,
    "Ares",       1,  1,  1,
    "Athena",     1,  1,  1,
    "Hephaestus", 1,  1,  1,
    "Aphrodite",  1,  1,  1,
    "Cyclope",    1,  1,  1,
    "Persephone", 1,  1,  1,
    "Hermes",     1,  1,  0,
    "Hebe",       1,  1,  0,
    "Dionysus",   1,  1,  0
  ) %>% 
  mutate(across(c(L, A, Y), as.factor))
```

Fit logistic regression model for outcome `Y`.

```{r}
fit <- glm(Y ~ A * L, family = binomial(), data = df)

broom::tidy(fit)
```

Estimate mean value of `Y` for each individual when untreated (`A` = 0) and treated (`A` = 1).

```{r}
pred_means <- 
  tibble(
    Y_A_0 = 
      mean(predict(fit, newdata = df %>% mutate(A = "0"), type = "response")),
    Y_A_1 = 
      mean(predict(fit, newdata = df %>% mutate(A = "1"), type = "response"))
  )

kable(pred_means)
```

The standardized mean outcomes were 0.50 for both the untreated and treated.

We'll now use the same procedure with the NHEFS data.

Fit linear regression model for weight gain.

```{r}
fit <- 
  lm(
    wt82_71 ~ qsmk * smokeintensity + sex + age + I(age^2) + race + education +
      wt71 + I(wt71^2) + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) +
      active + exercise,
    data = nhefs_censored
  )

broom::tidy(fit)
```

Estimate mean value of weight gain for each individual (`wt82_71`) when untreated (`qsmk` = 0) and treated (`qsmk` = 1).

```{r}
pred_means <- 
  tibble(
    wt82_71_qsmk_0 = mean(predict(fit, newdata = nhefs %>% mutate(qsmk = "0"))),
    wt82_71_qsmk_1 = mean(predict(fit, newdata = nhefs %>% mutate(qsmk = "1")))
  )

kable(pred_means)
```

The standardized mean in the untreated is `r format(pred_means$wt82_71_qsmk_0, digits = 2, nsmall = 2)` kg, and the standardized mean in the treated is `r format(pred_means$wt82_71_qsmk_1, digits = 2, nsmall = 2)` kg. So the estimate for the ATE is `r format(pred_means$wt82_71_qsmk_1 - pred_means$wt82_71_qsmk_0, digits = 1, nsmall = 1)` kg.

We will now use bootstrapping to obtain a confidence interval for this estimate.

ATE using fit on sample of data.

```{r}
ate <- function(data, sample_rows) {
  fit <- 
    lm(
      wt82_71 ~ qsmk * smokeintensity + sex + age + I(age^2) + race +
        education + wt71 + I(wt71^2) + I(smokeintensity^2) +
        smokeyrs + I(smokeyrs^2) + active + exercise,
      data = data %>% slice(sample_rows)
    )
  
  mean(predict(fit, newdata = nhefs %>% mutate(qsmk = "1"))) -
    mean(predict(fit, newdata = nhefs %>% mutate(qsmk = "0")))
}
```

Perform bootstrap resampling.

```{r}
set.seed(231)

n_boot <- 1e4

boot_out <- boot::boot(data = nhefs_censored, statistic = ate, R = n_boot)
```

Distribution of average treatment effect.

```{r}
tibble(ate = boot_out$t) %>% 
  ggplot(aes(ate)) +
  geom_histogram(binwidth = 0.1, boundary = 0) +
  labs(
    title = "Distribution of average treatment effect",
    x = "Average treatment effect",
    y = "Count"
  )
```

Estimate of ATE with 95% confidence interval calculated using bias-corrected and accelerated (BCa) method.

```{r}
estimate <- ate(data = nhefs_censored, sample_rows = 1:nrow(nhefs_censored))

v <- 
  broom::tidy(boot_out, conf.int = TRUE, conf.method = "bca") %>% 
  transmute(estimate, conf_low = conf.low, conf_high = conf.high)

kable(v, nsmall = 1)
```

```{r, echo=FALSE}
v %>% 
  save_result(
    file = "data/ate.rds",
    section = "13.3",
    method = "Standardization",
    dataset = "Full"
  )
```

