---
title: "16 Instrumental variable estimation"
author: "Bill Behrman"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Packages
library(tidyverse)

# Parameters
  # NHEFS data
file_nhefs <- here::here("data/nhefs.rds")
  # Common code
file_common <- here::here("_common.R")

# Print tibble
kable <- function(x, nsmall = 2, ...) {
  knitr::kable(x, format.args = list(digits = nsmall, nsmall = nsmall), ...)
}

#===============================================================================

# Run common code
source(file_common)

# NHEFS data with `qsmk` as a double
nhefs <- 
  read_rds(file_nhefs) %>% 
  mutate(qsmk = as.double(qsmk) - 1)

# NHEFS censored to remove missing weight and price measurements in 1982
nhefs_censored <- 
  nhefs %>% 
  drop_na(wt82, wt82_71, price82)
```

## 16.1 The three instrumental conditions

In the following with will consider `price82` as an instrumental variable:

* `price82`: Average tobacco price in state of residence in 1982 ($USD 2008)

Range of tobacco prices.

```{r}
summary(nhefs_censored$price82)
```

The prices ranged from \$`r format(min(nhefs_censored$price82), digits = 2, nsmall = 2)` to \$`r format(max(nhefs_censored$price82), digits = 2, nsmall = 2)` with an average of around \$`r format(mean(nhefs_censored$price82), digits = 2, nsmall = 2)`.

Distribution of tobacco prices.

```{r}
nhefs_censored %>% 
  ggplot(aes(price82)) + 
  geom_histogram(binwidth = 0.05, boundary = 0) +
  geom_vline(xintercept = mean(nhefs_censored$price82), color = "red") +
  labs(
    title = "Distribution of tobacco prices",
    subtitle = "Red line indicates mean price",
    x = "Price of tobacco",
    y = "Count"
  )
```

Proportion of subjects who quit smoking vs. mean price of tobacco.

```{r}
v <- 
  nhefs_censored %>% 
  arrange(price82, seqn) %>% 
  mutate(decile = cut_number(price82, n = 10)) %>% 
  group_by(decile) %>% 
  summarize(across(c(price82, qsmk), mean))

v %>% 
  ggplot(aes(price82, qsmk)) +
  geom_point() +
  scale_y_continuous(breaks = scales::breaks_width(0.02)) +
  labs(
    title = "Proportion of subjects who quit smoking vs. mean price of tobacco",
    subtitle = "By price decile",
    x = "Mean price of tobacco in decile",
    y = "Proportion of subjects who quit smoking in decile "
  )
```

By price decile, there is a very week association, if any, of the mean price of tobacco and the proportion of subjects who quit smoking.

Add variable `highprice` to indicate that `price82` is greater that 1.5.

```{r}
nhefs_censored <- 
  nhefs_censored %>% 
  mutate(
    highprice =
      case_when(
        price82 <= 1.5 ~ 0,
        price82 > 1.5 ~ 1,
        TRUE ~ NA_real_
      ) %>% 
      as.factor()
  )
```

Percentage of those who quit smoking by price group.

```{r}
v <- 
  nhefs_censored %>% 
  group_by(highprice) %>% 
  summarize(qsmk_pct = 100 * mean(qsmk))

kable(v, nsmall = 1)
```

`r format( v$qsmk_pct[v$highprice == "1"], digits = 1, nsmall = 1)`% of those in the `highprice` = 1 group quit smoking, and `r format(v$qsmk_pct[ v$highprice == "0"], digits = 1, nsmall = 1)`% of those in the `highprice` = 0 group quit smoking. The risk difference is therefore `r format(v$qsmk_pct[v$highprice == "1"] - v$qsmk_pct[v$highprice == "0"], digits = 1, nsmall = 1)`%.

## 16.2 The usual IV estimand

Mean values of weight gain (`wt82_71`) and whether subject quit smoking (`qsmk`) by price group.

```{r}
v <- 
  nhefs_censored %>% 
  group_by(highprice) %>% 
  summarize(across(c(wt82_71, qsmk), mean))

knitr::kable(v, digits = 4)
```

The differences in the means.

```{r}
v <- 
  v %>% 
  summarize(
    wt82_71_diff = wt82_71[highprice == "1"] - wt82_71[highprice == "0"],
    qsmk_diff = qsmk[highprice == "1"] - qsmk[highprice == "0"]
  )

knitr::kable(v, digits = 4)
```

Estimate of ATE.

```{r}
v %>% 
  summarize(estimate = wt82_71_diff / qsmk_diff) %>% 
  kable(nsmall = 1)
```

Equivalent estimate of ATE calculated using two saturated linear models.

```{r}
tibble(
  estimate =
    coef(lm(wt82_71 ~ highprice, data = nhefs_censored))["highprice1"] /
    coef(lm(qsmk ~ highprice, data = nhefs_censored))["highprice1"]
) %>% 
  kable(nsmall = 1)
```

Fit two-stage least-squares regression.

```{r}
fit <- ivreg::ivreg(wt82_71 ~ qsmk | highprice, data = nhefs_censored)

summary(fit)
```

Here's how to interpret the diagnostic tests:

* Weak instruments: A good instrumental variable is highly correlated with endogenous predictor variables and uncorrelated with the errors. This would be indicated with a high value for the statistic and a low p-value. Thus, `highprice` does not appear to be a good instrumental value.

* Wu-Hausman: Ordinary least squares can be inconsistent when predictor variables are correlated with the errors. This would be indicated with a high value for the statistic and a low p-value. This does not appear to be the case.

The F-statistic for the first-stage model is less than 10.

```{r}
lm(qsmk ~ highprice, data = nhefs_censored) %>% 
  broom::glance() %>% 
  transmute(f_statistic = statistic) %>% 
  kable(nsmall = 1)
```

Estimate of ATE with 95% confidence interval.

```{r}
broom::tidy(fit, conf.int = TRUE) %>%
  filter(term == "qsmk") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

The confidence interval is very large.

We will now estimate the ATE using g-estimation.

For potential counterfactual (`psi`), calculate logistic regression coefficient (`alpha`) and its p-value (`p_value`).

```{r}
g_est <- function(psi) {
  geepack::geeglm(
    highprice ~ h,
    family = binomial(),
    data = 
      nhefs_censored %>% 
      mutate(
        highprice = as.double(highprice) - 1,
        h = wt82_71 - psi * qsmk
      ),
    id = seqn
  ) %>% 
    broom::tidy() %>% 
    filter(term == "h") %>% 
    transmute(psi, alpha = estimate, p_value = p.value)
}
```

To estimate the ATE, use optimization to search for the value of `psi` with `alpha` value closest to 0.

```{r, warning=FALSE}
f <- function(psi) {
  abs(g_est(psi)$alpha)
}

v <- optimize(f, interval = c(-50, 50))

estimate <- v$minimum
tibble(estimate, `abs(alpha)` = v$objective) %>% 
  kable(nsmall = 3)
```

Thus the estimate of the ATE is consistent with the value above.

```{r}
tibble(estimate) %>% 
  kable(nsmall = 1)
```

The method of section 14.6 to find a confidence interval for the ATE does not work in this case, due to the inability of reducing the p-value to below 0.05.

## 16.5 The three instrumental conditions revisited

Use two-stage least-squares regression to calculate ATE with 95% confidence interval with `highprice` defined using `price`.

```{r}
ate <- function(price) {
  ivreg::ivreg(
    wt82_71 ~ qsmk | highprice,
    data = 
      nhefs_censored %>%
      mutate(
        highprice =
          case_when(
            price82 <= price ~ 0,
            price82 > price ~ 1,
            TRUE ~ NA_real_
          ) %>% 
          as.factor()
      )
  ) %>% 
    broom::tidy(conf.int = TRUE) %>%
    filter(term == "qsmk") %>% 
    select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
    add_column(price, .before = "estimate")
}
```

Estimates of ATEs with 95% confidence intervals for different prices.

```{r}
seq(1.6, 1.9, 0.1) %>% 
  map_dfr(ate) %>% 
  mutate(price = str_c("$", format(price, nsmall = 2))) %>% 
  kable(nsmall = 1)
```

The estimates, in kilograms, are too large to be plausible, and the confidence intervals are very large.

Fit two-stage least-squares regression using variables and model from section 15.1. We will assume that `qsmk` is the only endogenous predictor variable.

```{r}
fit <- 
  ivreg::ivreg(
    wt82_71 ~ sex + age + I(age^2) + race + education + wt71 + I(wt71^2) +
      smokeintensity + I(smokeintensity^2) + smokeyrs + I(smokeyrs^2) + active +
      exercise | qsmk | highprice,
    data = nhefs_censored
  )

summary(fit)
```

The weak instruments diagnostic test again indicates that `highprice` does not appear to be a good instrumental value.

Estimate of ATE with 95% confidence interval.

```{r}
broom::tidy(fit, conf.int = TRUE) %>%
  filter(term == "qsmk") %>% 
  select(estimate, conf_low = conf.low, conf_high = conf.high) %>% 
  kable(nsmall = 1)
```

The confidence interval is again very large.

